---
title: "Cross-checks"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cross-checks}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(PowerUpR)
```

To install `PowerUpR`:
```{r,  message=FALSE, eval=FALSE}
install.packages("PowerUpR")
```


### Individual-level Random Assignment Design

```{r, message=FALSE}
d1 <- mdes.ira1r1(power=.80, n=250)
power.ira1r1(es=d1$mdes[1], n=250)
mrss.ira1r1(es=d1$mdes[1])
```

### Two-Level Blocked (Constant Treatment Effect) Individual-level Random Assignment Design, Treatment at Level 1

```{r, message=FALSE}
d1 <- mdes.bira2c1(power=.80, n=15, J=20)
power.bira2c1(es=d1$mdes[1], n=15, J=20)
mrss.bira2c1(es=d1$mdes[1], n=15)
```


### Two-Level Blocked (Fixed Treatment Effect) Individual-level Random Assignment Design, Treatment at Level 1

```{r, message=FALSE}
d1 <- mdes.bira2f1(power=.80, n=15, J=20)
power.bira2f1(es=d1$mdes[1], n=15, J=20)
mrss.bira2f1(es=d1$mdes[1], n=15)
```

### Two-Level Blocked Individual-level Random Assignment Design, Treatment at Level 1

```{r, message=FALSE}
d1 <- mdes.bira2r1(power=.80, rho2=.17, omega2=.50, n=15, J=20)
power.bira2r1(es=d1$mdes[1], rho2=.17, omega2=.50, n=15, J=20)
mrss.bira2r1(es=d1$mdes[1], rho2=.17, omega2=.50, n=15)
```


### Two-Level Cluster-level Random Assignment Design, Treatment at Level 2

```{r, message=FALSE}
d1 <- mdes.cra2r2(power=.80, rho2=.17, n=15, J=20)
power.cra2r2(es=d1$mdes[1], rho2=.17, n=15, J=20)
mrss.cra2r2(es=d1$mdes[1], rho2=.17, n=15)
```

### Three-Level (Fixed Treatment Effect) Blocked Cluster-level Random Assignment Design, Treatment at Level 2

```{r, message=FALSE}
d1 <- mdes.bcra3f2(power=.80, rho2=.10, n=20, J=44, K=5)
power.bcra3f2(es=d1$mdes[1], rho2=.10, n=20, J=44, K=5)
mrss.bcra3f2(es=d1$mdes[1], rho2=.10, n=20, J=44)
```

### Three-Level Blocked Individual-level Random Assignment Design, Treatment at Level 1

```{r, message=FALSE}
d1 <- mdes.bira3r1(power=.80, rho3=.20, rho2=.15,
                   omega3=.10, omega2=.10,
                   n=69, J=10, K=100)
power.bira3r1(es=d1$mdes[1], rho3=.20, rho2=.15,
              omega3=.10, omega2=.10,
              n=69, J=10, K=100)
mrss.bira3r1(es =d1$mdes[1], rho3=.20, rho2=.15,
             omega3=.10, omega2=.10,
             n=69, J=10)
```

### Three-Level Blocked Cluster-level Random Assignment Design, Treatment at Level 2

```{r, message=FALSE}
d1 <- mdes.bcra3r2(power=.80, rho3=.13, rho2=.10, omega3=.4,
                   n=10, J=6, K=24)
power.bcra3r2(es=d1$mdes[1], rho3=.13, rho2=.10, omega3=.4,
              n=10, J=6, K=24)
mrss.bcra3r2(es=d1$mdes[1], rho3=.13, rho2=.10, omega3=.4,
             n=10, J=6)
```

### Three-Level Cluster-level Random Assignment Design, Treatment at Level 3

```{r, message=FALSE}
d1 <- mdes.cra3r3(power=.80, rho3=.06, rho2=.17, n=15, J=3, K=60)
power.cra3r3(es=d1$mdes[1], rho3=.06, rho2=.17, n=15, J=3, K=60)
mrss.cra3r3(es=d1$mdes[1], rho3=.06, rho2=.17, n=15, J=3)

```

### Four-Level (Fixed Treatment Effect) Blocked Cluster-level Random Assignment Design, Treatment at Level 3

```{r, message=FALSE}
d1 <- mdes.bcra4f3(power=.80, rho3=.15, rho2=.15,
                   n=10, J=4, K=4, L=15)
power.bcra4f3(es=d1$mdes[1], rho3=.15, rho2=.15,
              n=10, J=4, K=4, L=15)
mrss.bcra4f3(es=d1$mdes[1], rho3=.15, rho2=.15,
             n=10, J=4, K=4)
```

### Four-Level Blocked Individual-level Random Assignment Design, Treatment at Level 1

```{r, message=FALSE}
d1 <- mdes.bira4r1(power=.80, rho4=.05, rho3=.15, rho2=.15,
                   omega4=.50, omega3=.50, omega2=.50,
                   n=10, J=4, K=4, L=27)
power.bira4r1(es=d1$mdes[1], rho4=.05, rho3=.15, rho2=.15,
              omega4=.50, omega3=.50, omega2=.50,
              n=10, J=4, K=4, L=27)
mrss.bira4r1(es=d1$mdes[1], rho4=.05, rho3=.15, rho2=.15,
             omega4=.50, omega3=.50, omega2=.50,
             n=10, J=4, K=4)
```

### Four-Level Blocked Cluster-level Random Assignment Design, Treatment at Level 2

```{r, message=FALSE}
d1 <- mdes.bcra4r2(power=.80, rho4=.05, rho3=.15, rho2=.15,
                   omega4=.50, omega3=.50, n=10, J=4, K=4, L=20)
power.bcra4r2(es=d1$mdes[1], rho4=.05, rho3=.15, rho2=.15,
              omega4=.50, omega3=.50, n=10, J=4, K=4, L=20)
mrss.bcra4r2(es=d1$mdes[1], rho4=.05, rho3=.15, rho2=.15,
             omega4=.50, omega3=.50, n=10, J=4, K=4)
```

### Four-Level Blocked Cluster-level Random Assignment Design, Treatment at Level 3

```{r, message=FALSE}
d1 <- mdes.bcra4r3(power=.80, rho4=.05, rho3=.15, rho2=.15,
                   omega4=.50, n=10, J=4, K=4, L=20)
power.bcra4r3(es=d1$mdes[1], rho4=.05, rho3=.15, rho2=.15,
              omega4=.50, n=10, J=4, K=4, L=20)
mrss.bcra4r3(es=d1$mdes[1], rho4=.05, rho3=.15, rho2=.15,
             omega4=.50, n=10, J=4, K=4)
```

### Four-Level Cluster-level Random Assignment Design, Treatment at Level 4

```{r, message=FALSE}
d1 <- mdes.cra4r4(power=.80, rho4=.05, rho3=.05, rho2=.10,
                  n=10, J=2, K=3, L=20)
power.cra4r4(es=d1$mdes[1], rho4=.05, rho3=.05, rho2=.10,
             n=10, J=2, K=3, L=20)
mrss.cra4r4(es=d1$mdes[1], rho4=.05, rho3=.05, rho2=.10,
            n=10, J=2, K=3)
```
