---
title: "Vignette A: Designing a Three-Level Cluster Randomized Trial to Detect Main Treatment Effect"
date: "July 07, 2017"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette A - Main Effect}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

-----
*Note*: This vignette refers to `PowerUpR` development version (v0.1.3.9000). Regression discontinuity designs, moderator effects, unequal cost per unit, and bound constraints are not available in earlier versions (<= v0.1.3). In earlier versions COSA is supported in `optimal` functions. 
-----

# Problem

Institue of Education Science (IES) reported that evidence on effectiveness of Saxon Math curriclum on secondary school student's algebra subject is lacking and further research is needed (U.S. Department of Education, 2016). To evaluate the effectiveness Saxon Math curriculum, assume a 3-level cluster randomized trial is considered where schools are randomly assigned to treatment (Saxon Math curriclum) and a control condition. In the grant proposal number of schools, teachers, and students, and the cost associated with data collection from this sample needs to be justified. See Vignette B for designing a three-level cluster randomized trial that is sensetive to moderator effect. 

## Statistical Model
Due to nesting of students within classrooms and nesting of classrooms within schools, the treatment effect ($\delta$) can be estimated via the following 3-level hierarchical linear model (HLM, Raudenbush & Bryk, 2002)

$$\begin{eqnarray}
  L1: Y_{ijk} &=& \beta_{0jk} + \beta_{1jk}X + r_{ijk}, \quad r_{ijk} \thicksim N(0,\sigma_{|X}^2) \newline
  L2: \beta_{0jk} &=& \gamma_{00k} + \gamma_{01k}W + \mu_{0jk}, \quad \mu_{0jk} \thicksim N(0,\tau_{2|W}^2) \newline
      \beta_{1jk} &=& \gamma_{10k} \newline
  L3: \gamma_{00k} &=& \xi_{001} + \delta T_k + \xi_{002}V + \varsigma_{00k}, \quad \varsigma_{00k} \thicksim N(0,\tau_{3|V}^2) \newline
      \gamma_{01k} &=& \xi_{010} \newline
      \gamma_{10k} &=& \xi_{100}
  \end{eqnarray}$$

where, $Y$: Student algebra posttest score. $X$: Student algebra pretest score. $W$: Classroom mean algebra pretest score. $V$: School mean algebra pretest score. $T$: Treatment status. Consider the following hypothetical values for design parameters

- Proportion of variance in algebra posttest at the school level ($\tau_3^2$): $.15$
- Proportion of variance in algebra posttest at the classroom level ($\tau_2^2$):  $.05$
- Proportion of variance in algebra posttest at the student level ($\sigma^2$): $.65$
- Proportion of variance in algebra posttest explained by school mean algebra pretest: ($R_3^2 = 1 - \frac{\tau_{3|V}^2}{\tau_{3}^2}$):  $.45$
- Proportion of variance in algebra posttest explained by classroom mean algebra pretest: ($R_2^2 = 1 - \frac{\tau_{2|W}^2}{\tau_{2}^2}$): $.50$
- Proportion of variance in algebra posttest explained by student algebra pretest: ($R_1^2 = 1 - \frac{\sigma_{|X}^2}{\sigma^2}$): $.55$

Then, intraclass correlation coefficients are

  - $\rho_2 = \tau_2^2 / (\sigma^2 + \tau_2^2 + \tau_3^2) = .05 /(.65 + .15 + .05) = 0.059$ 
  - $\rho_3 = \tau_3^2 / (\sigma^2 + \tau_2^2 + \tau_3^2) = .15 /(.65 + .15 + .05) = 0.176$ 

## Minimum Detectable Effect Size (MDES)
### Scenario 1: Find MDES
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Number of schools ($K$): $83$
- Average number of classrooms per school ($J$): $2$
- Average number of students per classroom ($n$) : $10$
- Statistical power ($1-\beta$): $.80$ 

Given design parameters and further assumptions, what is minimum detectable effect size?
```{r, message=FALSE, warning=FALSE}
library(PowerUpR)
```

```{r, message=FALSE}
design1 <- mdes.cra3r3(power=.80, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40, n=10, J=2, K=83)
```

## Statistical Power
### Scenario 2: Find Statistical Power
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Number of schools ($K$): $60$
- Average number of classrooms per school ($J$): $2$
- Average number of students per classroom ($n$) : $10$
- Minimum detectable effect size (MDES): $.232$

Given design parameters and further assumptions, what is statistical power?

```{r, message=FALSE}
design2 <- power.cra3r3(mdes=.23, rho2=.06, rho3=.18,
                        g3=1, R12=.55, R22=.50, R32=.45,
                        P=.40, n=10, J=2, K=83)
```


## Constrained Optimal Sample Allocation

### Scenario 3: Number of Schools (K)
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Average number of classrooms per school ($J$): $2$
- Average number of students per classroom ($n$) : $10$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.232$

Given design parameters and further assumptions, how many schools should be recruited?

```{r, message=FALSE}
design3 <- cosa.cra3r3(power=.80, mdes=.23, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40, n=10, J=2)
```

### Scenario 4: Number of Classrooms (J)
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Number of schools ($K$): $60$
- Average number of students per classroom ($n$) : $10$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.232$

Given design parameters and further assumptions, how many classrooms should be recruited per school on average?

```{r, message=FALSE}
design4 <- cosa.cra3r3(power=.80, mdes=.23, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40, n=10, K=83)
```

### Scenario 5: Number of Students (n)
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Number of schools ($K$): $60$
- Average number of classrooms per school ($J$): $2$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.232$
  
Given design parameters and further assumptions, how many students should be recruited per classroom on average?

```{r, message=FALSE}
design5 <- cosa.cra3r3(power=.80, mdes=.23, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40, J=2, K=83)
```

### Scenario 6: Constrain Power
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.232$
- Marginal cost per student ($cn$): $10$
- Marginal cost per classroom ($cJ$): $200$
- Marginal cost per school ($cK$): $500$
  
Given design parameters and further assumptions, how many schools, classrooms and students should be recruited to achieve 80% power while minimizing the total cost?

```{r, message=FALSE}
design6 <- cosa.cra3r3(cn=10, cJ=200, cK=500,
                       constrain="power", power=.80,
                       mdes=.23, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40)
```

Note that MDES values in the output are calculated at the specified power value (.80), and power values in the output are
calculated at the specified MDES value (.23). 

### Scenario 7: Constrain Power and Number of Classrooms
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.232$
- Average number of classrooms per school ($J$): $2$
- Marginal cost per student ($cn$): $10$
- Marginal cost per classroom ($cJ$): $200$
- Marginal cost per school ($cK$): $500$
  
Given design parameters and further assumptions, how many schools, and students should be recrutied to achieve 80% power while minimizing the total cost?

```{r, message=FALSE}
design7 <- cosa.cra3r3(cn=10, cJ=200, cK=500,
                       constrain="power", power=.80, J=2, 
                       mdes=.23, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40)
```

Note that MDES values in the output are calculated at the specified power value (.80), and power values in the output are
calculated at the specified MDES value (.23). 

### Scenario 8: Constrain MDES and Number of Classrooms
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.20$
- Average number of classrooms per school ($J$): $2$
- Marginal cost per student ($cn$): $10$
- Marginal cost per classroom ($cJ$): $200$
- Marginal cost per school ($cK$): $500$
  
Given design parameters and further assumptions, how many schools, and students should be recruited to achieve an MDES of .20 while minimizing the total cost?

```{r, message=FALSE}
design8 <- cosa.cra3r3(cn=10, cJ=200, cK=500,
                       constrain="mdes", mdes=.20, J=2,
                       power=.80, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40)
```

Note that MDES values in the output are calculated at the specified power value (.80), and power values in the output are
calculated at the specified MDES value (.20). 

### Scenario 9: Constrain Cost and Number of Classrooms
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.20$
- Average number of classrooms per school ($J$): $2$
- Marginal cost per student ($cn$): $10$
- Marginal cost per classroom ($cJ$): $200$
- Marginal cost per school ($cK$): $500$
- Budget: $130,000$
  
Given design parameters and further assumptions, how many schools, and students should be recruited while maximizing the power?

```{r, message=FALSE}
design9 <- cosa.cra3r3(cn=10, cJ=200, cK=500,
                       constrain="cost", cost=130000, J=2,
                       power=.80, mdes=.20, rho2=.06, rho3=.18,
                       g3=1, R12=.55, R22=.50, R32=.45,
                       P=.40)
```

Note that MDES values in the output are calculated at the specified power value (.80), and power values in the output are
calculated at the specified MDES value (.20).  

### Scenario 10: Constrain Cost, Number of Classrooms, and Number of Students
Further assumptions 

- Alpha level ($\alpha$): $0.05$
- Statistical power ($1-\beta$): $.80$ 
- Minimum detectable effect size (MDES): $.20$
- Average number of students per classroom ($n$): $20$
- Average number of classrooms per school ($J$): $2$
- Marginal cost per student ($cn$): $10$
- Marginal cost per classroom ($cJ$): $200$
- Marginal cost per school ($cK$): $500$
- Budget: $130,000$
  
Given design parameters and further assumptions, how many schools should be recruited while maximizing the power?

```{r, message=FALSE}
design10 <- cosa.cra3r3(cn=10, cJ=200, cK=500,
                        constrain="cost", cost=130000, n=20, J=2,
                        power=.80, mdes=.20, rho2=.06, rho3=.18,
                        g3=1, R12=.55, R22=.50, R32=.45,
                        P=.40)
```

Note that MDES values in the output are calculated at the specified power value (.80), and power values in the output are
calculated at the specified MDES value (.20). 

## Summary
In reality, the cost associated with recruting treatment and control schools may not be equal, and consequently, we may not want to fix proportion of treatment schools ($P$). How many schools do we need and how much would that cost, if control schools cost half as much. 
```{r, message=FALSE}
design11 <- cosa.cra3r3(cn=c(10,5), cJ=c(200,100), cK=c(500,250),
                        constrain="cost", cost=130000, n=20, J=2, ncase=5,
                        power=.80, mdes=.20, rho2=.06, rho3=.18,
                        g3=1, R12=.55, R22=.50, R32=.45,
                        P=.40)
## intgeger approximations
design11$integer.optim
## get confidence interval for MDES
mdes95CI <- cosa.to.mdes(design11)
## get percentile equivalance
mdes.to.pctl(mdes95CI)
```

A 3-level hierarchical linear model is proposed to estimate treatment effect, where schools are randomly assigned to treatment and control conditions with 40% probability. We assume that proportion of variance in algebra posttest explained by school mean algebra pretest is 45%, proportion of variance in algebra posttest explained by classroom mean algebra pretest is 50%, proportion of variance in algebra posttest explained by student algebra pretest is 55%. Furthermore we assume intraclass correlation coefficient at the school level is 17.6%, and intraclass correlation coefficient at the classroom level is 5.9%. 

Having \$10 marginal cost per student, \$200 marginal cost per classroom, \$500 marginal cost per school in treatment group, and half as much cost per student, per classroom and per school in control group, based on `PowerUpR` package (Bulus, Dong, Kelcey, Spybrook, & Le, 2017), with a budget of \$130000 we can afford recruting 143 schools in total, each with two classrooms and 40 students on average. For a two-tailed hypothesis test for treatment effect, and Type I error rate of 5%, such sample is estimated to have statistical power of 91% for an effect size of .169 95% CI[.050, .287], which is equaivalent to increasing an average student's algebra posttest score from 50th percentile to 57th percentile. 

## Relevant Citations

Bulus, M., & Dong, N., Kelcey, B., Spybrook, J., Le, B. (2017). `PowerUpR`: R version of PowerUp!. R package version 1.1.3.

Dong, N., & Maynard, R. A. (2013a). PowerUp!: A Tool for Calculating Minimum Detectable Effect Sizes and Minimum Required Sample Sizes
for Experimental and Quasi-Experimental Design Studies, *Journal of Research on Educational Effectiveness, 6(1)*, 24-6.

Dong, N., & Maynard, R. A. (2013b). PowerUp!: A tool for calculating minimum detectable
effect sizes and minimum required sample sizes for experimental and quasi-experimental
design studies. [Software]. Accessed from [http://www.causalevaluation.org/](http://www.causalevaluation.org/).

U.S. Department of Education, Institute of Education Sciences, What Works Clearinghouse. (2016, May).  Secondary Mathematics intervention report: Saxon Math. Retrieved from [https://ies.ed.gov/ncee/wwc/Docs/InterventionReports/wwc_saxon_052416.pdf](https://ies.ed.gov/ncee/wwc/Docs/InterventionReports/wwc_saxon_052416.pdf)

